Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2016-07-06 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Started on Ch 5 CF Comments

As defined in NFPA 921, a fire is the visible or measurable physical changes, or identifiable shapes, formed by a fire effect or group of fire effects~\cite{NFPA_921}.  In order to measure the dimensions of the fire pattern, such as height and width, a determination had to be made as to where the fire pattern stops.  The boundary between the undamaged area and the area that has been damaged by the fire is known as the line of demarcation~\cite{NFPA_921}.  As presented in the introduction, DeHaan~\cite{DeHaan:2012} categorizes the fire effects that form patterns as follows:
\begin{enumerate}
	\item Surface deposits -– no irreversible effect on surface
	\item Surface thermal effects –- physical change such a discoloration or melting
	\item Charring -– evidence of surface burning
	\item Penetration -– charring below the surface
	\item Consumption –- loss of surface material, charring throughout
\end{enumerate}

@@

Soot and discoloration alone was not used in this study, since the gasoline and the polyurethane foam generate significant quantities of soot as show in Figure~\ref{Soot_pattern}.  From the photograph the areas of penetration and consumption can be clearly delineated while the discolored and sooted areas tend to meld together.  Further the focus of the measurements in this study are thermal, therefore the focus is on the burned, not burned line of demarcation.

@@

\chapter{Bench Scale Experiments}

@@

The cone calorimeter is one of the most commonly used bench-scale rate of heat release apparatus that uses the oxygen consumption method. It was developed at NIST in the 1980s~\cite{babrauskas:1984}. The cone calorimeter has been adopted as ASTM~E1354, Test Method for Heat and Visible Smoke Release Rates for Materials and Products Using an Oxygen Consumption Calorimeter~\cite{ASTM_E1354}. The cone calorimeter consists of a heater, spark ignitor, sample holder, and load cell located underneath an exhaust hood. Typically, the sample is located in the open with free access of air to the combustion zone. The heater consists of a 5~kW electrical heating element inside an insulated stainless-steel conical shell. Samples can be tested in either a horizontal or vertical orientation. When tests are performed in the horizontal configuration, the specimen is positioned approximately 25~mm beneath the bottom plate of the cone heater. Flames and products of combustion pass through a circular opening at the top of the heater. The heater can expose samples to a maximum irradiance of approximately 100~kW/m$^2$. Figure~\ref{Cone_Cal} is a schematic of the cone calorimeter.

@@

To conduct the experiments, the radiant panel was ignited and allowed to pre-heat for at least 60 minutes.  The water cooled, Schmidt-Boelter total heat flux gauge was mounted in the center the sample holder, and the heat flux positions were confirmed. The heat flux gauge was then removed from the sample holder trolley. Just before the sample was mounted on the movable trolley, an aluminum heat shield was placed between the radiant panel and the sample holder trolley.  Once the sample was mounted and the sample was moved to the appropriate distance from the radiant panel based on the heat flux desired for the experiment, the heat shield was removed.  The removal of the heat shield would be recorded as the beginning of the exposure and therefore the beginning of the experiment.   Each sample was exposed for 600 seconds. Each sample had a single bare-bead, Chromel-Alumel (type K) thermocouple, with a 0.5~mm nominal diameter bead in contact with the center point of each sample. During the experiments the samples were monitored for temperature and ignition. After the exposure time was completed the heat shield was once again placed between the sample and the radiant panel.  The samples were cooled, removed and examined.  The area of thermal damage to the painted paper surface of the gypsum board was compared with exposure heat flux.  This was another means of potentially obtaining a measure of the critical heat flux.

@@

The results from the three different bench scale experiments are presented and discussed in this section.  All three test methods provided some insight into the ignition and the critical heat flux required for ignition or flame spread of the painted paper surface on the gypsum wall board.  The cone calorimeter also provided values for heat release rate.

@@

Additional cone calorimeter experiments were conducted in the University of Canterbury Fire Laboratory.  In these experiments the purpose was to obtain an estimate of the temperature of the painted paper surface at the time of ignition.   Samples of painted gypsum wallboard were shipped from the NIST Fire Research Laboratory in Gaithersburg, MD., USA.  In these experiments an additional incident heat flux level of 30~kg/m$^2$ was added to provide another test between the 25 and 35~kW/m$^2$ levels.  The averaged ignition time and temperature pairs are shown in Table~\ref{tab:Gypsum_wallboard_igntemp}.  The average time to ignition values between the two different calorimeters are within $\pm~50~\%$ and $\pm~10~\%$ at incident heat flux values of 35 and 50~kg/m$^2$, respectively.  The average ignition temperatures from the nine experiments ranged from 335 to 375~$^{\circ}$C.

>>>>

As defined in NFPA 921, a fire pattern is the visible or measurable physical changes, or identifiable shapes, formed by a fire effect or group of fire effects~\cite{NFPA_921}.  In order to measure the dimensions of the fire pattern, such as height and width, a determination had to be made as to where the fire pattern stops.  The boundary between the undamaged area and the area that has been damaged by the fire is known as the line of demarcation~\cite{NFPA_921}.  As presented in the introduction, DeHaan~\cite{DeHaan:2012} categorizes the fire effects that form patterns as follows:
\begin{enumerate}
	\item Surface deposits -– no irreversible effect on surface
	\item Surface thermal effects –- physical change such a discoloration or melting
	\item Charring -– evidence of surface burning
	\item Penetration -– charring below the surface
	\item Consumption –- loss of surface material, charring throughout
\end{enumerate}

@@

Soot and discoloration alone was not used in this study, since the gasoline and the polyurethane foam generate significant quantities of soot as shown in Figure~\ref{Soot_pattern}.  From the photograph the areas of penetration and consumption can be clearly delineated while the discolored and sooted areas tend to meld together.  Further the focus of the measurements in this study are thermal, therefore the focus is on the burned, not burned line of demarcation.

@@

\chapter{Bench Scale Assessment of Thermal Damage to Gypsum Wallboard}

@@

The cone calorimeter is one of the most commonly used bench-scale rate of heat release apparatus that uses the oxygen consumption calorimetry method. It was developed at NIST in the 1980s~\cite{babrauskas:1984}. The cone calorimeter has been adopted as ASTM~E1354, Test Method for Heat and Visible Smoke Release Rates for Materials and Products Using an Oxygen Consumption Calorimeter~\cite{ASTM_E1354}. The cone calorimeter consists of a conical shaped heater, spark ignitor, sample holder, and load cell located underneath an exhaust hood. Typically, the sample is located in the open with free access of air to the combustion zone. The heater consists of a 5~kW electrical heating element inside an insulated stainless-steel conical shell. Samples can be tested in either a horizontal or vertical orientation. When tests are performed in the horizontal configuration, the specimen is positioned approximately 25~mm beneath the bottom plate of the cone heater. Flames and products of combustion pass through a circular opening at the top of the heater. The heater can expose samples to a maximum irradiance of approximately 100~kW/m$^2$. Figure~\ref{Cone_Cal} is a schematic of the cone calorimeter.

@@

To conduct the experiments, the radiant panel was ignited and allowed to pre-heat for at least 60 minutes.  The water cooled, Schmidt-Boelter total heat flux gauge was mounted in the center the sample holder, and the heat flux positions were confirmed. The heat flux gauge was then removed from the sample holder trolley. Just before the sample was mounted on the movable trolley, an aluminum heat shield was placed between the radiant panel and the sample holder trolley.  Once the sample was mounted and the sample was moved to the appropriate distance from the radiant panel based on the heat flux desired for the experiment, the heat shield was removed.  The removal of the heat shield was recorded as the beginning of the exposure and therefore the beginning of the experiment.   Each sample was exposed for 600 seconds. Each sample had a single bare-bead, Chromel-Alumel (type K) thermocouple, with a 0.5~mm nominal diameter bead in contact with the center point of each sample. During the experiments the samples were monitored for temperature and ignition. After the exposure time was completed the heat shield was once again placed between the sample and the radiant panel.  The samples were cooled, removed and examined.  The area of thermal damage to the painted paper surface of the gypsum board was compared with exposure heat flux.  This was another means of potentially obtaining a measure of the critical heat flux.

@@

The results from the three different bench scale experiments are presented and discussed in this section.  All three test methods provided some insight into the ignition and the critical heat flux required for ignition or flame spread of the painted paper surface on the gypsum wallboard.  The cone calorimeter also provided values for heat release rate.

@@

Additional cone calorimeter experiments were conducted in the University of Canterbury Fire Laboratory.  In these experiments the purpose was to obtain an estimate of the temperature of the painted paper surface at the time of ignition.   Samples of painted gypsum wallboard were shipped from the NIST Fire Research Laboratory in Gaithersburg, MD., USA.  In these experiments an additional incident heat flux level of 30~kW/m$^2$ was added to provide another test between the 25 and 35~kW/m$^2$ levels.  The averaged ignition time and temperature pairs are shown in Table~\ref{tab:Gypsum_wallboard_igntemp}.  The average time to ignition values between the two different calorimeters are within $\pm~50~\%$ and $\pm~10~\%$ at incident heat flux values of 35 and 50~kW/m$^2$, respectively.  The average ignition temperatures from the nine experiments ranged from 335 to 375~$^{\circ}$C.
